---
title: "Council accounts"
output: html_notebook
---

# Council accounts - combining horizontally

This notebook details the process of combining council accounts.

The data comes from the Ministry of Housing, Communities & Local Government's [Local authority revenue expenditure and financing collection](https://www.gov.uk/government/collections/local-authority-revenue-expenditure-and-financing). 

We have downloaded the data for 2010/11 onwards with the intention of using the script to add new figures when new data is published.

## Breaking down the problem

Here are the tasks we need to complete:

1. Combine the spreadsheets
2. Perform basic cleaning (remove redundant rows, combine/clarify headings across multiple cells)
3. Create a subset containing the data we want to focus on
4. Perform analysis

The files are in the same folder in this script with names like *RS_2015-16_data_by_LA_-_Revision.xlsx* - but 5 use one format and the latest 2 use a different format.

Let's start with activating some useful packages.

```{r}
#Activate the library for handling Excel files, and for calling URLs
library(readxl)
library(httr)
```

## Clean the spreadsheet while importing

Now to import those sheets, and clean along the way. 

There are two problems here: first, the column headers don't start until row 4; and second, those headers are spread across 3 rows.

The most detailed headings are in row 7, with related codes for each heading in row 6, and broader categories in row 5. So we skip to row 7 for the main dataset, and store the other headings in another data frame just in case.

This is best stored in a function so we can run it on the new data, but we test it first.

```{r}
#Read that file into a new data frame
#We want the 3rd sheet, and to skip the first 6 lines so row 7 is used for headers
#We also ask it to class dashes and lack of data as NA
ro1516 <- read_excel("RS_2015-16_data_by_LA_-_Revision.xlsx", 
                              sheet = 3, 
                              skip = 6, 
                              na = "c('-','')",
                     col_types = "guess") #This helps numeric columns be treated as such
summary(ro1516)
#Grab the other headings
ro1516.otherheads <- read_excel("RS_2015-16_data_by_LA_-_Revision.xlsx", 
                                sheet = 3, skip = 4, n_max = 1, range = "A6:IR6")


ro1617 <- read_excel("RS_2016-17_data_by_LA.xlsx", 
                              sheet = 3, 
                              skip = 6, 
                              na = "c('-','')")
# Test the columns are the same
colnames(ro1516) == colnames(ro1617)

```

The columns are not the same. So we need to extract a few and create two data frames which are the same:

```{r}
#Check the colnames
colnames(ro1516) 
#Add to the front of each the code from the other list
colnames(ro1516) <- paste(head(colnames(ro1516.otherheads),length(colnames(ro1516))),
                          colnames(ro1516), sep=":")
colnames(ro1516) 
```


## Create a subset

There are 252 variables in this spreadsheet, so let's narrow down to the ones we want to focus on.

We can use regex to find the columns where 'Social Care' is mentioned:

```{r}
#The grepl function is used to generate a series of TRUE/FALSE matches for each column
#That is then nested within square brackets after the colnames function to show those that are TRUE
colnames(ro1516)[grepl("Social Care",colnames(ro1516))]
#Let's try some others
colnames(ro1516)[grepl(".*Highway.*",colnames(ro1516))]
colnames(ro1516)[grepl(".*GFRA*",colnames(ro1516))]
colnames(ro1516)[grepl("Public Health",colnames(ro1516))]
colnames(ro1516)[grepl(".*Cultural*",colnames(ro1516))]
colnames(ro1516)[grepl(".*Environmental*",colnames(ro1516))]
colnames(ro1516)[grepl(".*Planning*",colnames(ro1516))]
colnames(ro1516)[grepl(".*Education*",colnames(ro1516))]
#And reserves
colnames(ro1516)[grepl(".*at 1 April*",colnames(ro1516))]
```

We also need to grab the first few columns:

```{r}
ro1516.las <- ro1516[c(1,2,3,5)]
```



Now that we think we have the right columns, let's generate that subset:

```{r}
listofterms <- "Social Care|Highway|GFRA|Public Health|Cultural|Environmental|Planning|Education|at 1 April"
#We remove 'colnames' now so we are accessing the data frame as a whole
ro1516.sub <- ro1516[grepl(listofterms,
                           colnames(ro1516))]
#Add in the LA details extracted earlier
ro1516.sub <- cbind(ro1516.las,ro1516.sub)
```

We've now gone from 252 variables to 20, which is going to be much easier to deal with.

## Storing in a function

The question is, can we store all those steps in a function so we don't have to write them all over again?

```{r}
stripbackro.horiz <- function(filename, year){
  #read from Excel into an object 
  roobject <- read_excel(filename, 
                          sheet = 3, 
                          skip = 6, 
                          na = "c('-','')",
                         col_types = "guess")
  #Store the first few columns separately
  roobject.las <- roobject[c(1,2,3,5)]
  #Read from Excel again to grab the codes
  roobject.otherheads <- read_excel(filename, 
                                sheet = 3, skip = 4, n_max = 1, range = "A6:IR6")
  #Combine the codes with the relevant column headings
  colnames(roobject) <- paste(head(colnames(roobject.otherheads),length(colnames(roobject))),
                          colnames(roobject), sep=": ")
  #Add the year as a prefix
  colnames(roobject) <- paste(year, colnames(roobject), sep=": ")
  #Create a list of terms that we can use to filter
  listofterms <- "Social Care|Highway|GFRA|Public Health|Cultural|Environmental|Planning|Education|at 1 April"
  #Use grepl to filter to just those columns containing the keywords
  roobject.sub <- roobject[grepl(listofterms,
                           colnames(roobject))]
  #Add in the LA details extracted earlier
  roobject.sub <- cbind(roobject.las,roobject.sub)
  #No need to add the year in this version
  #roobject.sub$year <- year
  #Return to whatever called this function
  return(roobject.sub)
}
```

And can we remove the 'other' and shire districts..?

### Subsetting by type of organisation

We also need to remove all the organisations we don't want. The type of organisation is shown in the 'Class' column:

```{r}
table(ro1516$`X__1:Class`)
```

We don't want shire districts (SD), or other authorities such as police or fire (O). 

```{r}
stripbackro.horiz <- function(filename, year){
  #read from Excel into an object 
  roobject <- read_excel(filename, 
                          sheet = 3, 
                          skip = 6, 
                          na = "c('-','')",
                         col_types = "guess")
  #Store the first few columns separately
  roobject.las <- roobject[c(1,2,3,5)]
  #Read from Excel again to grab the codes
  roobject.otherheads <- read_excel(filename, 
                                sheet = 3, skip = 4, n_max = 1, range = "A6:IR6")
  #Combine the codes with the relevant column headings
  colnames(roobject) <- paste(head(colnames(roobject.otherheads),length(colnames(roobject))),
                          colnames(roobject), sep=": ")
  #Add the year as a prefix
  colnames(roobject) <- paste(year, colnames(roobject), sep=": ")
  #Create a list of terms that we can use to filter
  listofterms <- "Social Care|Highway|GFRA|Public Health|Cultural|Environmental|Planning|Education|at 1 April"
  #Use grepl to filter to just those columns containing the keywords
  roobject.sub <- roobject[grepl(listofterms,
                           colnames(roobject))]
  #Add in the LA details extracted earlier
  roobject.sub <- cbind(roobject.las,roobject.sub)
  #No need to add the year in this version
  #roobject.sub$year <- year
  #Subset so we don't have those with Class codes O or SD
  roobject.sub <- subset(roobject.sub,roobject.sub$Class != "O" & roobject.sub$Class != "SD")
  #Subset so we only have rows where the code is not NA
  roobject.sub <- subset(roobject.sub,roobject.sub$`E-code` != "NA")
  #Return to whatever called this function
  return(roobject.sub)
}
```

And let's try it:

```{r}
#Supply 2 arguments: the filename and the year
ro1617.sub.horiz <- stripbackro.horiz("RS_2016-17_data_by_LA.xlsx", 2017)
#Show first few rows
head(ro1617.sub.horiz)
table(ro1617.sub.horiz$Class)
```


```{r}
#Supply 2 arguments: the filename and the year
ro1516.sub.horiz <- stripbackro.horiz("RS_2015-16_data_by_LA_-_Revision.xlsx", 2016)
#Show first few rows
head(ro1516.sub.horiz)
table(ro1516.sub.horiz$Class)
```

## Combining the files

We can try a `cbind` 

```{r}
ro.all.horiz <- cbind(ro1516.sub.horiz,ro1617.sub.horiz)
```

This will only work if both have the same number of rows, and also assumes that the rows are in the same order. We can re-do this with a proper `join` function using the `sqldf` package later.


## Repeating for the older datasets

The older datasets have a different structure so need to be imported and handled separately before being combined.

```{r}
#Read the file into a new data frame
#We want the 3rd sheet, and to skip the first 5 lines so row 6 is used for headers
#We also ask it to class dashes and lack of data as NA
ro1011 <- read_excel("Revenue_Outturn_Summary__RS__data_2010-11_by_LA_-_27-Nov-2012-v2.xls", sheet = 3, skip = 5, na = "c('-','')",col_types = "guess")

ro1112 <- read_excel("Revenue_Outturn_Summary__RS__data_2011-12_by_LA_-_Revised_28-Nov-2013.xls", sheet = 3, skip = 5, na = "c('-','')",col_types = "guess")

ro1213 <- read_excel("Revenue_Outturn_Summary__RS__data_2012-13_by_LA__Revised__-_18-Feb-2014.xls", sheet = 3, skip = 5, na = "c('-','')",col_types = "guess")
```

Again we can see that the column names will need to be made consistent. 

First, though, let's create a function to whittle the data down

Once again, use regex to find the columns where 'Social Care' and other terms are mentioned:

```{r}
#The grepl function is used to generate a series of TRUE/FALSE matches for each column
#That is then nested within square brackets after the colnames function to show those that are TRUE
colnames(ro1011)[grepl(".*Education*",colnames(ro1011))]
colnames(ro1011)[grepl(".*Highway.*",colnames(ro1011))]
colnames(ro1011)[grepl("Social Care",colnames(ro1516))]
colnames(ro1011)[grepl("Public Health",colnames(ro1011))]
colnames(ro1011)[grepl(".*GFRA*",colnames(ro1011))]
colnames(ro1011)[grepl(".*Cultural*",colnames(ro1011))]
colnames(ro1011)[grepl(".*Environmental*",colnames(ro1011))]
colnames(ro1011)[grepl(".*Planning*",colnames(ro1011))]
#And reserves - this doesn't work with 1 April so we are more literal
colnames(ro1011)[grepl(".*reserves*",colnames(ro1011))]
```

Some of these are the fields we want:

```{r}
colnames(ro1516.sub)
```




Looking at the spreadsheets themselves we can note that most of the key fields are in the first 12 columns, but as the years go on, child social care is added in 2011/12 and then public health in 2012/13. In 2014/15 everything shifts one column to the right as they add region, and ONS codes appear. That dataset might need to be handled separately.

It is tricky anyway: we need to account for columns of missing data.

```{r}
stripbackropre14.horiz <- function(filename, year){
  #read from Excel into an object 
  roobject <- read_excel(filename, 
                          sheet = 3, 
                          skip = 5, 
                          na = "c('-','')",
                         col_types = "guess")
  #Store the first 14 columns separately - these have most info
  roobject.sub <- roobject[seq(1,14)]
  #Use grepl to filter to just those columns containing the keyword 'reserves'
  roobject.res <- roobject[grepl("reserves",
                           colnames(roobject))]
  #Add to other object
  roobject.sub <- cbind(roobject.sub,roobject.res)
  #Subset so we don't have those with Class codes O or SD
  roobject.sub <- subset(roobject.sub,roobject.sub$Class != "O" & roobject.sub$Class != "SD")
  #Subset so we only have rows where the code is not NA
  roobject.sub <- subset(roobject.sub,roobject.sub$`E-code` != "NA")
  #Add the year as a prefix
  colnames(roobject.sub) <- paste(year, colnames(roobject.sub), sep=": ")
  #Return to whatever called this function
  return(roobject.sub)
}
```

Now test:

```{r}
ro1011.sub.horiz <- stripbackropre14.horiz("Revenue_Outturn_Summary__RS__data_2010-11_by_LA_-_27-Nov-2012-v2.xls",2011)

ro1112.sub.horiz <- stripbackropre14.horiz("Revenue_Outturn_Summary__RS__data_2011-12_by_LA_-_Revised_28-Nov-2013.xls",2012)

ro1213.sub.horiz <- stripbackropre14.horiz("Revenue_Outturn_Summary__RS__data_2012-13_by_LA__Revised__-_18-Feb-2014.xls",2013)

ro1314.sub.horiz <- stripbackropre14.horiz("Revenue_Outturn_Summary__RS__data_2013-14_by_LA_-_Revised_28-Jan-2015.xls",2014)

ro1415.sub.horiz <- stripbackropre14.horiz("Revenue_Outturn_Summary__RS__data_2014-15_by_LA_-_02-Feb-2016.xls",2015)
```

And then combine:

```{r}
ro.all.horiz <- cbind(ro1011.sub.horiz,ro1112.sub.horiz,ro1213.sub.horiz,ro1314.sub.horiz,ro1415.sub.horiz,ro1516.sub.horiz,ro1617.sub.horiz)
colnames(ro.all.horiz)
```


## Cleaning data types

To check if we need more cleaning we can generate a summary:

```{r}
summary(ro.all.horiz)
```

## Showing changes

```{r}
#First make a copy to work on
ro.all.change <- ro.all.horiz
#Create a check column
ro.all.change$does11match16 <- ro.all.change$`2011: E-code` == ro.all.change$`E-code`
table(ro.all.change$does11match16)
#This line works for 2016 but 2017 throws an error as it is stored as character
ro.all.change$highways11to16 <- ro.all.change$`2016: 290: Highways and transport services` - ro.all.change$`2011: Highways and transport services`
head(ro.all.change$highways11to16)
```

Here's a problem column:

```{r}
summary(ro.all.change$`2017: 290: Highways and transport services`)
```

Let's check it converts to numeric OK:


```{r}
head(ro.all.change$`2017: 290: Highways and transport services`)
head(as.numeric(ro.all.change$`2017: 290: Highways and transport services`))
```
And use that to calculate a difference:

```{r}
ro.all.change$highways11to17 <- as.numeric(ro.all.change$`2017: 290: Highways and transport services`) - ro.all.change$`2011: Highways and transport services`
head(ro.all.change$highways11to17)
```

## Try SQL

Test with a simple one first:

```{r}
library(sqldf)
sqldf("select Class, count(*) from ro1011 
      group by Class
      order by count(*) desc")
```

You can perform calculations:

```{r}
sqldf("select `Local authority`, `Other services` - `Central services` as otherminuscentral from ro1011
      order by otherminuscentral desc")
```


```{r}
#sqldf won't work with ro.all.change or ro.all.horiz
#but for some reason recreating it this way solves it
roallhoriz <- ro.all.horiz[seq(1,144)]
sqldf("select count(*) from roallhoriz")
head(roallhoriz$`2011: Cultural and related services`)
head(roallhoriz$`2017: 509: Cultural and related services`)
changeculture11to17 <- sqldf("select `Local authority`, (`2017: 509: Cultural and related services` * 1000) - (`2011: Cultural and related services` * 1000) as changeculture11to17, `2017: 509: Cultural and related services` * 1000, `2011: Cultural and related services` * 1000 from roallhoriz
      order by changeculture11to17 asc")
changeculture11to17
```

Let's try some others:

```{r}
changeenviro11to17 <- sqldf("select `Local authority`, (`2017: 590: Environmental and regulatory services` * 1000) - (`2011: Environmental and regulatory services` * 1000) as changeculture11to17, (`2017: 590: Environmental and regulatory services` * 1000 - `2011: Environmental and regulatory services` * 1000) / (`2011: Environmental and regulatory services` * 1000) as changeperc, `2017: 590: Environmental and regulatory services` * 1000, `2011: Environmental and regulatory services` * 1000 from roallhoriz
      order by changeperc asc")
mean(changeenviro11to17$changeperc)
median(changeenviro11to17$changeperc)
changeenviro11to17
```
Now planning

```{r}
#This worked below so something must have been changed in the lines below which now breaks this
changeplanning11to17 <- sqldf("select `Local authority`, (
`2017: 599: Planning and development services` * 1000) - (
`2011: Planning and development services` * 1000) as changeculture11to17, (
`2017: 599: Planning and development services` * 1000 - 
`2011: Planning and development services` * 1000) / (
`2011: Planning and development services` * 1000) as changeperc, 
`2017: 599: Planning and development services` * 1000, 
`2011: Planning and development services` * 1000 from roallhoriz
      order by changeperc asc")
mean(changeplanning11to17$changeperc)
median(changeplanning11to17$changeperc)
changeplanning11to17
```

We can't go back to 2010/11 for the change in social care because the figures were combined then.

```{r}

changechildsoc12to17 <- sqldf("select `Local authority`, (
`2017: 330: Children Social Care` * 1000) - (
`2012: Children Social Care` * 1000) as changeculture11to17, (
`2017: 330: Children Social Care` * 1000 - 
`2012: Children Social Care` * 1000) / (
`2012: Children Social Care` * 1000) as changeperc, 
`2017: 330: Children Social Care` * 1000, 
`2012: Children Social Care` * 1000 from roallhoriz
      order by changeperc asc")
mean(changechildsoc12to17$changeperc)
median(changechildsoc12to17$changeperc)
changechildsoc12to17
```

We can't go back to 2010/11 for the change in social care because the figures were combined then.

```{r}
changeotherearmarked11to17 <- sqldf("select `Local authority`, (`2017: 915: Other earmarked financial reserves level at 1 April` * 1000) - (`2011: Other earmarked financial reserves level` * 1000) as changeculture11to17, (`2017: 915: Other earmarked financial reserves level at 1 April` * 1000 - `2011: Other earmarked financial reserves level` * 1000) / (`2011: Other earmarked financial reserves level` * 1000) as changeperc, `2017: 915: Other earmarked financial reserves level at 1 April` * 1000, `2011: Other earmarked financial reserves level` * 1000 from roallhoriz order by changeperc asc")
mean(changeotherearmarked11to17$changeperc)
median(changeotherearmarked11to17$changeperc)
changeotherearmarked11to17
```

## SQL queries across multiple objects

Instead of querying the combined data we could use SQL to query the earliest and latest datasets.

```{r}
sqldf("SELECT ro1011.`E-code`, ro1011.`Local authority`, ro1011.`Class`, ro1011.`Cultural and related services` AS culture1011, ro1617.`Cultural and related services` as culture1617, ro1617.`Cultural and related services` - ro1011.`Cultural and related services` AS culchange1117, (ro1617.`Cultural and related services` - ro1011.`Cultural and related services`)/ro1011.`Cultural and related services` AS culpercchange1117 from ro1011 
      LEFT JOIN ro1617 
      ON ro1011.`Local authority` = ro1617.`Local authority`
      WHERE ro1011.`Class` != 'O' AND ro1011.`Class` != 'SD' AND ro1011.`E-code` != 'NA'
      ORDER BY culpercchange1117 ASC
      ")
```

