---
title: "Council accounts"
output: html_notebook
---

# Council accounts - analysis

With the accounts downloaded, cleaned, and filtered into a subset (see other notebook in this folder), this notebook deals with the analysis.

## Create a working copy

We need to create a working copy first so the original is untouched.

```{r}
#Create a working copy
reservesdataonly.councils.wc <- reservesdataonly.councils
```

## Convert characters to numbers

The `tidyverse` package can help us clean up data types:

```{r}
library(tidyverse)
#Generate a summary of what it would look like if we parsed as double (number with decimals)
summary(parse_double(reservesdataonly.councils.wc[,5]))
#Now replace that column with a version parsed as integer
reservesdataonly.councils.wc[,5] <- parse_double(reservesdataonly.councils.wc[,5])
#Summarise that column 
summary(reservesdataonly.councils.wc[,5])
```

We need to repeat this for any other columns we want to work with.

```{r}
colnames(reservesdataonly)
colnames(reservesdataonly.headings)
```


## Totals spent

We can calculate a basic total for a column:

```{r}
table(reservesdataonly.councils.wc[,5])
sum(reservesdataonly.councils.wc[,5])
```

But the column needs to be numeric first:

```{r}
sum(reservesdataonly.councils.wc[,6])
```

```{r}
#Parse as doubles
parse_double(reservesdataonly.councils.wc[,6])
#Replace column with numeric version
reservesdataonly.councils.wc[,6] <- parse_double(reservesdataonly.councils.wc[,6])
#Sum - this returns NA
sum(reservesdataonly.councils.wc[,6])
#Table to show frequency of numbers
table(reservesdataonly.councils.wc[,6])
```

## Comparing two columns for different years

Let's compare the unallocated financial reserves for 2017 and 2018. First, find the columns:

```{r}
colnames(reservesdataonly.councils.wc)
```

Columns 15 and 19 seem to be the key ones:

```{r}
#colnames(reservesdataonly.councils)[grepl(".*unallocated.*",colnames(reservesdataonly.councils))]
colnames(reservesdataonly.councils.wc[15])
colnames(reservesdataonly.councils.wc[19])
colnames(reservesdataonly.councils.wc[14])
colnames(reservesdataonly.councils.wc[18])
```

As before, these are probably stored as character, so let's check that and then convert to number:

```{r}
reservesdataonly.councils.wc[,15]
summary(reservesdataonly.councils.wc[,15])
reservesdataonly.councils.wc[,15] <- parse_number(reservesdataonly.councils.wc[,15])
summary(reservesdataonly.councils.wc[,15])
```

Then repeat for column 19:

```{r}
reservesdataonly.councils.wc[,19]
summary(reservesdataonly.councils.wc[,19])
reservesdataonly.councils.wc[,19] <- parse_number(reservesdataonly.councils.wc[,19])
summary(reservesdataonly.councils.wc[,19])
```


Then repeat for column 14:

```{r}
reservesdataonly.councils.wc[,14]
summary(reservesdataonly.councils.wc[,14])
reservesdataonly.councils.wc[,14] <- parse_number(reservesdataonly.councils.wc[,14])
summary(reservesdataonly.councils.wc[,14])
```

And column 18

```{r}
reservesdataonly.councils.wc[,18]
summary(reservesdataonly.councils.wc[,18])
reservesdataonly.councils.wc[,18] <- parse_number(reservesdataonly.councils.wc[,18])
summary(reservesdataonly.councils.wc[,18])
```

We are getting errors for the same rows, so we need to strip those out before we can do a calculation.

But because we will need to reintegrate later when exporting, we need to store it first:

```{r}
#First store the NA values
reservesdataonly.councils.naonly <- subset(reservesdataonly.councils.wc,is.na(reservesdataonly.councils.wc[,18]))
#Then remove them so we can do analysis
reservesdataonly.councils.wc <- subset(reservesdataonly.councils.wc,!is.na(reservesdataonly.councils.wc[,18]))
```



Let's create a column showing the amount spent from unallocated reserves (positive numbers = spending):


```{r}
#parse_number used in this and below calculations as parse_integer generates NA results where there are decimal places
reservesdataonly.councils.wc$unallocated_reserves_spend <- parse_number(reservesdataonly.councils.wc[,15])-parse_number(reservesdataonly.councils.wc[,19])
```

And now one showing the spending in earmarked reserves:

```{r}
reservesdataonly.councils.wc$earmarked_reserves_spend <- parse_number(reservesdataonly.councils.wc[,14])-parse_number(reservesdataonly.councils.wc[,18])
```

Now a column showing the total in reserves, allocated and earmarked at the start of the year

```{r}
reservesdataonly.councils.wc$totalreserves_april <- parse_number(reservesdataonly.councils.wc[,14])+parse_number(reservesdataonly.councils.wc[,15])
```

Now a column showing the total in reserves, allocated and earmarked at the end of the year
```{r}
reservesdataonly.councils.wc$totalreserves_march <- parse_number(reservesdataonly.councils.wc[,18])+parse_number(reservesdataonly.councils.wc[,19])
```

Now a column showing the total spent from reserves, allocated and earmarked over the financial year

```{r}
#Because the number of columns may vary I've specified the index of the last, and the penultimate (ncol-1) columns instead of 28 and 29
reservesdataonly.councils.wc$totalreservesspend <- parse_number(reservesdataonly.councils.wc[,ncol(reservesdataonly.councils.wc)-1])-parse_number(reservesdataonly.councils.wc[,ncol(reservesdataonly.councils.wc)])
```


## Recombine and export

Now we can bring in those NA rows - but we need to give it the same extra columns so it will fit:

```{r}
reservesdataonly.councils.naonly$unallocated_reserves_spend <- ""
reservesdataonly.councils.naonly$totalreserves_april <- ""
reservesdataonly.councils.naonly$totalreservesspend <- ""
```

Then combine using `rbind`:

```{r}
reservesdataonly.councils.export <- rbind(reservesdataonly.councils.wc, reservesdataonly.councils.naonly)
```

And export

```{r}
write.csv(reservesdataonly.councils.export,"councilaccountsanalysis.csv")
```



